Does an earthquake “know” how big it will be? A neural-net aided study

### 2024-07-30 article.tex
Final version before submission

\title[Running Title]{Does an earthquake “know” how big it will be? A neural-net aided study} % 2nd option: Using past seismicity to predict the magnitude of future earthquakes

\author[1,2]{\fnm{Neri} \sur{Berman}}\email{neriberman@gmail.com}
\author[2]{\fnm{Oleg} \sur{Zlydenko}}
\author[2]{\fnm{Oren} \sur{Gilon}}
\author[1]{\fnm{Yohai} \sur{Bar-Sinai}}\email{ybarsinai@gmail.com}

@@

\abstract{
Earthquake occurrence is notoriously difficult to predict. While some aspects of their spatiotemporal statistics can be relatively well captured by point-process models, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts. This is due both to the lack of information about fault conditions and to the inherent complexity of rupture dynamics. Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events besides the time-independent Gutenberg Richter (GR) distribution, which describes the marginal distribution over large regions and long times. This approach implicitly assumes that earthquake magnitudes are independent of seismic history and are identically distributed. In this work we challenge this view by showing that information about the magnitude of an upcoming earthquake can be directly extracted from the seismic history. We present MAGNET - MAGnitude Neural EsTimation model, an open-source, geophysically-inspired neural-network model for probabilistic forecasting of future magnitudes from cataloged properties: hypocenter locations, occurrence times and magnitudes of past earthquakes. Our history-dependent model outperforms stationary and quasi-stationary state of the art GR-based benchmarks, in real catalogs in Southern California, Japan and New-Zealand.  This demonstrates that earthquake catalogs contain information about the magnitude of future earthquakes, prior to their occurrence. We conclude by proposing methods to apply the model in characterization of the preparatory phase of earthquakes, and in operational hazard alert and earthquake forecasting systems.
}

@@

To this end, we construct MAGnitude Neural EsTimation model, \nobreak MAGNET, a neural-based model that predicts the magnitude of a given earthquake given the short and long term seismic history prior to its occurrence. Our model is trained on seismic catalogs containing information about past earthquakes, including location, time, and magnitude. In some experiments, we also incorporate focal mechanism information. Importantly, the model is not tasked with predicting the timing and location of the event, as they are explicitly provided. Thus, we separate out the task of modeling nucleation statistics and isolate the question of magnitude predictability. If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of a specific earthquake is extractable from cataloged properties alone.

@@

In addition to comparing the log-likelihood of the observed data, one can convert the prediction problem to a binary question: will the earthquake at the spacetime query $\textbf{x}_i, t_i$ have a higher magnitude than a threshold $m_t$? This can be easily calculated from the prediction, as the probability that $m_i>m_t$ is simply $\int_{m_t}^{\infty}p(m')dm'$.
This task, asking whether a given event will be a large earthquake, is more interpretable and can be evaluated using standard binary classification metrics such as area under the receiver operator (ROC) curve \cite{Murphy} and under the interpolated precision-recall (PR) curve \cite{buttcher_information_2010} which suit better an imbalanced data set such as ours. Fig.~\ref{fig:metrics}d-f and g-i show these metrics for all three regions examined and compare against the GR-variant models. Again, MAGNET outperforms all benchmarks.

@@

One such artifact is ``short-time incompleteness'' (STI): the fact that immediately following a large event the signal of subsequent small events might be buried in the coda of the mainshock~\cite{kagan_short-term_2004, stockman_forecasting_2023}. Thus, immediately after a large event it is less likely that a small event will be included in the catalog, not because it did not occur, but rather because it was not detected. Indeed, our model consistently allocates less probability mass to low magnitudes shortly after large events (see the Tohoku aftershock in Fig.~\ref{fig:intro_fig}b), and some of the information gain may be attributed to this effect.

@@

The spatially conditioned CIG are shown by the red dashed curves in Fig. \ref{fig:info_gain_over_time}. It is a steadily increasing function for the Southern California and Japan data sets. In the New Zealand data set the curve shows an initial increase followed by a plateau after the first sequence of major events (the 2016 Te Araroa and Kaikōura earthquakes). This may indicate that a good portion of this information gain followed by these earthquakes does originate from spatially inhomogeneity, as also hinted by the spatially-varying GR benchmark. Nevertheless, the overall score still shows an advantage for our model over the test set.
Fig. \ref{fig:info_gain_over_time}e-g shows the spatial distribution of MAGNET's advantage over the GR benchmark. All three regions exhibit a spatially diffuse pattern, suggesting that information is not gained solely from any one specific area.

\section{Implications and Outlook}
In this work we have used neural based models to find statistical dependencies between earthquake magnitudes and seismic history. Since our model is provided with the known location and timing of an event and is only tasked with predicting its magnitude, we have separated out the question of rate and location of earthquakes, and directly probe whether their magnitude is history dependent. MAGNET shows robust and significant information gain, even after accounting for possible measurement artifacts such as temporal or spatially dependent properties compared against a host of benchmarks, thereby answering an open question in earthquake predictability. Our results indicate that some information about the magnitude of a specific event can be extracted, on average, from the regional seismic history in the form of a simple hypo-center catalog.

@@

Understanding such patterns will advance the effort of earthquake prediction a great deal\cite{mousavi_deep-learning_2022, mousavi_machine_2023, mignan_neural_2020, karimpouli_explainable_2023, bergen_machine_2019}, though it is still further down the road.
However, even as a ``black box'', such a model may be integrated into operational systems such early warning systems, which already provide a quick estimation of the source time and location. 
Another possible application is incorporating a history-dependent magnitude distribution into point-process rate forecasting models, such as the epidemic type aftershock sequence model (ETAS) or similar \cite{ogata_statistical_1988, ogata_statistics_2017, dascher-cousineau_using_2023}. As discussed above, such models currently assume no statistical dependence of the magnitude. Since larger earthquakes produce more aftershocks, a more accurate estimation of magnitudes will have a positive compounding effect on the resulting statistics.

@@

\section*{Author Contributions}
N.B., O.Z., O.G. and Y.B.S. designed the research. N.B. conducted the experiments. N.B., O.G. and Y.B.S. preformed the statistical analysis. N.B. and O.Z. wrote the code base. N.B. and Y.B.S. wrote the manuscript. All authors reviewed the manuscript.

@@

>>>>

\title[Running Title]{Does an earthquake “know” how big it will be? A neural-net aided study}

\author[1,2]{\fnm{Neri} \sur{Berman}}\email{neriberman@gmail.com}
\author[2]{\fnm{Oleg} \sur{Zlydenko}}
\author[2]{\fnm{Oren} \sur{Gilon}}
\author[2]{\fnm{Yossi} \sur{Matias}}
\author[1]{\fnm{Yohai} \sur{Bar-Sinai}}\email{ybarsinai@gmail.com}

@@

\abstract{
Earthquake occurrence is notoriously difficult to predict. While some aspects of their spatiotemporal statistics can be relatively well captured by point-process models, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts. This is due both to the lack of information about fault conditions and to the inherent complexity of rupture dynamics. Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events besides the time-independent Gutenberg Richter (GR) distribution, which describes the marginal distribution over large regions and long times. This approach implicitly assumes that earthquake magnitudes are independent of previous seismicity and are identically distributed. In this work we challenge this view by showing that information about the magnitude of an upcoming earthquake can be directly extracted from the seismic history. We present MAGNET - MAGnitude Neural EsTimation model, an open-source, geophysically-inspired neural-network model for probabilistic forecasting of future magnitudes from cataloged properties: hypocenter locations, occurrence times and magnitudes of past earthquakes. Our history-dependent model outperforms stationary and quasi-stationary state of the art GR-based benchmarks, in real catalogs in Southern California, Japan and New-Zealand.  This demonstrates that earthquake catalogs contain information about the magnitude of future earthquakes, prior to their occurrence. We conclude by proposing methods to apply the model in characterization of the preparatory phase of earthquakes, and in operational hazard alert and earthquake forecasting systems.
}

@@

To this end, we construct MAGnitude Neural EsTimation model, \nobreak MAGNET, a neural-based model that predicts the magnitude of a given earthquake given the short and long term seismic history prior to its occurrence. Our model is trained on seismic catalogs containing information about past earthquakes, including location, time, and magnitude. In some experiments, we also incorporate focal mechanism information.
Importantly, the model is not tasked with predicting the timing and location of the event, as they are explicitly provided.
Thus, we separate out the task of modeling nucleation statistics and isolate the question of magnitude predictability.
If our model performs better than a draw from the GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of a specific earthquake is extractable from cataloged properties alone. 
An example of MAGNET's performance on a few select earthquakes is shown in Fig \ref{fig:intro_fig}a. For instance, MAGNET improved the likelihood of estimating the true magnitude for the Ridgecrest mainshock in Southern California and the Kaikōura event in New Zealand by factors of approximately 13 and 4, respectively, over the naive GR predictor.

@@

In addition to comparing the log-likelihood of the observed data, one can convert the prediction problem to a binary question: will the earthquake at the spacetime query $\textbf{x}_i, t_i$ have a higher magnitude than a threshold $m_t$? This can be easily calculated from the prediction, as the probability that $m_i>m_t$ is simply $\int_{m_t}^{\infty}p(m')dm'$.
This task, asking whether a given event will be a large earthquake, is more interpretable and can be evaluated using standard binary classification metrics such as area under the receiver operator (ROC) curve \cite{Murphy} and under the interpolated precision-recall (PR) curve \cite{buttcher_information_2010}. Fig.~\ref{fig:metrics}d-f and g-i show these metrics for all three regions examined and compare against the GR-variant models. Again, MAGNET outperforms all benchmarks.

@@

One such artifact is ``short-time incompleteness'' (STI): the fact that immediately following a large event the signal of subsequent small events might be buried in the coda of the mainshock~\cite{kagan_short-term_2004, stockman_forecasting_2023}. Thus, immediately after a large event, smaller events are less likely to be included in the catalog, not because they did not occur, but rather because they were not detected. Indeed, our model consistently allocates less probability mass to low magnitudes shortly after large events (see the Tohoku aftershock in Fig.~\ref{fig:intro_fig}b), and some of the information gain may be attributed to this effect.

@@

The spatially conditioned CIG are shown by the red dashed curves in Fig. \ref{fig:info_gain_over_time}. It is a steadily increasing function for the Southern California and Japan data sets. In the New Zealand data set the curve shows an initial increase followed by a plateau after the first sequence of major events (the 2016 Te Araroa and Kaikōura earthquakes). This may indicate that a good portion of this information gain followed by these earthquakes does originate from spatially inhomogeneity, as also hinted by the spatially-varying GR benchmark. Nevertheless, the overall score still shows an advantage for our model over the test set.
Fig. \ref{fig:info_gain_over_time}e-g shows the spatial distribution of MAGNET's advantage over the GR benchmark. All three regions exhibit a spatially diffuse pattern, suggesting that information is not gained solely from any one specific area (compare EM Fig.~\ref{fig:seismicity_grid} for the distribution of seismicity).

\section{Implications and Outlook}
In this work we have used neural based models to find statistical dependencies between earthquake magnitudes and seismic history. Since our model is provided with the known location and timing of an event and is only tasked with predicting its magnitude, we have separated out the question of rate and location of earthquakes, and directly probe whether their magnitude is history dependent. MAGNET shows robust and significant information gain, even after accounting for possible measurement artifacts such as temporal or spatially dependent properties compared against a host of benchmarks, thereby answering an open question in earthquake predictability. Our results indicate that some information about the magnitude of a specific event can be extracted, on average, from the regional seismic history in the form of a hypo-center catalog.

@@

Understanding such patterns will advance the effort of earthquake prediction a great deal\cite{mousavi_deep-learning_2022, mousavi_machine_2023, mignan_neural_2020, karimpouli_explainable_2023, bergen_machine_2019}, though it is still further down the road.
Additionally, such models may be incorporated in early warning systems, which already provide a quick (though not accurate) estimation of the source time and location. 
Another possible application is incorporating a history-dependent magnitude distribution into point-process rate forecasting models, such as the epidemic type aftershock sequence model (ETAS) or similar \cite{ogata_statistical_1988, ogata_statistics_2017, dascher-cousineau_using_2023}. As discussed above, such models currently assume no statistical dependence of the magnitude. Since larger earthquakes produce more aftershocks, a more accurate estimation of magnitudes will have a positive compounding effect on the resulting statistics.

@@

\section*{Author Contributions}
N.B., O.Z., O.G., Y.M. and Y.B.S. designed the research. N.B. conducted the experiments. N.B., O.G. and Y.B.S. preformed the statistical analysis. N.B. and O.Z. wrote the code base. N.B. and Y.B.S. wrote the manuscript. All authors reviewed the manuscript.

@@

\newpage
\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/seismicity_maps.pdf}
    \caption{
    \textbf{Event count in examined regions}. Number of events per bin is indicated in the adjacent color bar per inset. Bin divisions are identical to those presented in Figures \ref{fig:info_gain_over_time}e-g.
    }
    \label{fig:seismicity_grid}
\end{figure}
